Sources of Measurement Bias in Modern CPUs

### 2013-12-21 paper.tex
Backup

\documentclass[a4paper,11pt,twocolumn,twoside]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{listings}
\usepackage{authblk}

@@

\author{Lars Kirkholt Melhus}
\author{Rune Erlend Jensen}
\affil{Norwegian University of Science and Technology}
\title{Sources of Measurement Bias on Modern CPUs}
\date{} % This is a timeless piece

@@

In the paper appropriately named “Producing Wrong Data Without Doing Anything Obviously Wrong” by T. Mytkowicz et al. \cite{Mytkowicz:2009:WrongData}, the authors show how simple changes to the execution environment can have an impact on performance. 
One of their examples is a program where performance varies significantly with changes to the environment variables. 
Unix environment variables contain various information about the system, such as the name of the currently logged in user, home- and current directory path.
This means that the length of your name -- by extension you user login -- could in principle be the tipping point between good and bad performance.

Previous work focuses mostly on how to mitigate effects of bias, and avoid drawing the wrong conclusions based on misleading measurements \cite{Mytkowicz:2008:Easy, Mytkowicz:2009:WrongData}.
The goal of this paper is to show not only that bias exists, but also \emph{why} it occurs.

@@

Cache conflicts can in many cases be a potential explanation for bias effects observed by changing memory layout.
Although cache is often a likely explanation of performance variations, bias caused by cache issues is well understood and \emph{not} the focus of this paper.
Cache related metrics are monitored in order to rule out cache as the underlying cause of bias. 
The hit rates of micro-ops for each level of cache can be monitored by performance counters \cite{Intel:2012:OptimizationManual}.

\section{4K Address Aliasing}
An event known as “4K aliasing” can occur when the addresses of a store instruction followed by a load instruction differ by a multiple of 4 KiB.
On recent Intel architectures, only the last 12 address bits are used to determine if operations refer to the same address \cite{Intel:2012:OptimizationManual}.
This is a heuristic employed by the CPU to enable out of order execution of memory access operations.
Independent pairs of loads and stores where the last address bits differ can then be reordered and issued out of order.
Doing only a partial compare, \emph{false} dependencies can be detected when the addresses are equal $\bmod 4096$, as in the following example: 
\begin{lstlisting}[language={[x86masm]Assembler}]
mov %rax ($0x601010)
mov ($0x604010) %rbx
\end{lstlisting}
Situations like these causes the pipeline to stall and wait for the operations to finish in proper order.
The number of times this happens can be counted by the following performance counter:
\begin{description}
  \item{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}\marginpar{\small{Evnt Mask \\ 0x07 0x01}} \hfill \\
  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.
\end{description}
Performance implications of 4K aliasing is discussed to some extent in the optimization manual from Intel. 
There are even some concrete suggestions for how avoid aliasing cases to improve performance (Rule 8 and 9) \cite{Intel:2012:OptimizationManual}.
In the following sections, we look at how address aliasing can explain some instances of measurement bias.

\subsection{Bias from Environment Size}
Variations in performance with bias towards certain environment sizes has been observed by other authors.
Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack. 
Figure \ref{fig:stack} shows the relative positions of some important sections of memory at run time, assuming a 64 bit process mapped to virtual memory.

\pgfplotstableread{resources/stack-offset.dat}{\stackoffsettable}
\begin{figure*}[t]
  \caption{Bias from environment size}
  \begin{tikzpicture}
    \begin{axis}[
        width = \textwidth,
        height = 6cm,
        font = \small,
        xlabel=Bytes added to environment,
        domain = 0:8192,
        xtick = {0,1024,...,8192},
        xmin = 0,
        xmax = 8192,
        legend pos = outer north east,
        cycle list name = exotic
      ]
      %\draw [<->, line width=0.8, color=red] (axis cs:0,8e5) -- (axis cs:4096,8e5) ;
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
      %\addplot table[x expr = \thisrowno{0}*16, y = r0107:u ] \stackoffsettable ;
      \addlegendentry{cycles:u} ;
      \addlegendentry{r0107:u} ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

@@

\begin{figure}
  \begin{tikzpicture}[font=\footnotesize]

@@

  \end{tikzpicture}
  \caption{Important sections of virtual memory on a 64 bit process}
  \label{fig:stack}
\end{figure}

@@

To illustrate how this can cause bias, we revisit the example first presented in \cite{Mytkowicz:2009:WrongData}.
Their program contains five variables, two of which are stack allocated and i, j and k that are statically allocated.

\begin{lstlisting}[language=C]
static int i, j, k;
int main() {
  int g = 0, inc = 1;
  for(; g < 65536; g++) {
    i += inc;
    j += inc;
    k += inc; 
  }
  return 0;
}
\end{lstlisting}

Measuring all available performance counters for this program under different environment sizes.

\subsection{Heap Address Aliasing}
Address aliasing can be caused by conflicting pairs of load/store operations to any part of memory.
The previous section looked specifically at an example of collision between stack variables and static data.
Another scenario to consider is collisions in heap allocated memory.
In particular, code that operate on pairs of contiguous arrays can be vulnerable to 4K aliasing.
Acquiring dynamic memory at run time is usually done by calling malloc, which takes a number of bytes to allocate and returns a pointer to that area.
A typical malloc implementation uses two different strategies for how to allocate memory.
The \emph{heap} area in virtual memory is used for smaller allocations.
Programs are initially given a small heap area, which the allocator can increase by calling brk or sbrk.
For larger allocations, it is typically faster to call mmap instead of increasing the ``normal'' heap.
The heap starts at a low address close to the data section and grows towards higher addresses. 
Mmap allocates much closer to the stack.
While heap addresses can look like 0x16e30a0 or 0x1723020, pointers returned by mmap can for example be 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, mmap has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{Libc's version of malloc adds 16 bytes of medatada at the beginning, thus every mmap'ed address ends with 0x010}

Consider the function shown in Listing [alg:convolution].
It computes the convolution between an input array and a fixed five-element kernel, writing the result to another array.
For simplicity, endpoints are not handled.

\begin{lstlisting}[language=C]
static float k[5]={0.1,0.25,0.3,0.25,0.1};
void conv(int n, float *in, float *out)
{
  int i, j;
  for (i = 2; i < n-2; ++i) {
    out[i] = 0;
    for (j = 0; j < 5; ++j)
      out[i] += in[i-2+j] * k[j];
  }
}
\end{lstlisting}

\pgfplotstableread{bin/convolution.dat}{\convolutiontable}
\begin{figure}[t]
  \caption{Offseting arrays}
  \begin{tikzpicture}
    \begin{axis}[
        font = \small,
        xlabel=Number of float (4B) offset,
        cycle list name=black white
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convolutiontable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convolutiontable ;
      \addlegendentry{cycles:u} ;
      \addlegendentry{r0107:u} ;
    \end{axis}
  \end{tikzpicture}
\end{figure}

@@

Some effort has been made to model properties of the LSD on previous Intel architectures, as an aid in low-level optimization. 
Hundt et. al. \cite{Hundt:2011:MAO} looks at assembly rewrite rules for optimizing agains the Core 2 LSD.
For loops to be recognized, they have to fit within a set of restrictions imposed by the hardware implementation that tries to detect them. 
For instance, because the LSD cache is of fixed size there will be a maximum number of instructions a loop can contain for it to be recognized.
The capabilities of the LSD has improved with newer micro-architectures.
Some important limitations for Haswell's LSD are listed below.
\begin{itemize}
  \item Up to 12 chunk fetches of 32 B instruction memory.\footnote{The optimization manual actually states 8 chunk fetches for HTT enabled, and 11 for HTT disabled. Our experiments with a minimal test case shows that the LSD is active for loops spanning up to 12 chunks -- regardless of hyper threading.}
  \item Up to 28 micro-operations for HTT enabled, 56 for HTT disabled.
  \item Can not contain CALL or RET instructions
  \item Stack operations must match, i.e push must be matched with a pop.
\end{itemize}

@@

\onecolumn

>>>>

\documentclass[a4paper,11pt,twocolumn,twoside]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{listings}
\usepackage{authblk}
\usepackage{caption} % For margin top bottom

@@

\author{Lars Kirkholt Melhus}
\author{Rune Erlend Jensen}
\affil{Norwegian University of Science and Technology}
\title{Sources of Measurement Bias in Modern CPUs}
\date{} % This is a timeless piece

@@

In the paper appropriately named “Producing Wrong Data Without Doing Anything Obviously Wrong” by T. Mytkowicz et al.~\cite{Mytkowicz:2009:WrongData}, the authors show how simple changes to the execution environment can have an impact on performance. 
One of their examples is a program where performance varies significantly with changes to the environment variables. 
Unix environment variables contain various information about the system, such as the name of the currently logged in user, home- and current directory path.
This means that the length of your name -- by extension you user login -- could in principle be the tipping point between good and bad performance.

Previous work focuses mostly on how to mitigate effects of bias, and avoid drawing the wrong conclusions based on misleading measurements~\cite{Mytkowicz:2008:Easy, Mytkowicz:2009:WrongData}.
The goal of this paper is to show not only that bias exists, but also \emph{why} it occurs.

@@

Cache conflicts can in many cases be a potential explanation for bias effects observed by changing memory layout.
Although cache is often a likely explanation of performance variations, bias caused by cache issues is well understood and \emph{not} the focus of this paper.
Cache related metrics are monitored in order to rule out cache as the underlying cause of bias. 
The hit rates of micro-ops for each level of cache can be monitored by performance counters~\cite{Intel:2012:OptimizationManual}.

\section{4K Address Aliasing}
An event known as “4K aliasing” can occur when the addresses of a store instruction followed by a load instruction differ by a multiple of 4 KiB.
On recent Intel architectures, only the last 12 address bits are used to determine if operations refer to the same address \cite{Intel:2012:OptimizationManual}.
This is a heuristic employed by the CPU to enable out of order execution of memory access operations.
Independent pairs of loads and stores where the last address bits differ can then be reordered and issued out of order.
Doing only a partial compare, \emph{false} dependencies can be detected when the addresses are equal $\bmod 4096$, as in the following example: 
\begin{lstlisting}[language={[x86masm]Assembler}]
mov %rax ($0x601010)
mov ($0x604010) %rbx
\end{lstlisting}
Situations like these causes the pipeline to stall and wait for the operations to finish in proper order.
The number of times this happens can be counted by the following performance counter:
\begin{description}
  \item{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}\marginpar{\small{Evnt Mask \\ 0x07 0x01}} \hfill \\
  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.
\end{description}
Performance implications of 4K aliasing is discussed to some extent in the optimization manual from Intel. 
There are even some concrete suggestions for how avoid aliasing cases to improve performance (Rule 8 and 9)~\cite{Intel:2012:OptimizationManual}.
In the following sections, we look at how address aliasing can explain some instances of measurement bias.

\subsection{Bias from Environment Size}
Variations in performance with bias towards certain environment sizes has been observed by other authors.
Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack. 
Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at run time, assuming a 64 bit process mapped to virtual memory.

\pgfplotstableread{resources/stack-offset.dat}{\stackoffsettable}
\begin{figure*}[t]
  \caption{Bias from environment size for program in Listing~\ref{lst:loopkernel}. Measure of average 100 cycle counts for 512 different environments, spread over two 4K periods. Compiled with no optimization.}
  \label{fig:envbias}
  \begin{tikzpicture}
    \begin{axis}[
        width = \textwidth,
        height = 6cm,
        font = \small,
        xlabel=Bytes added to environment,
        domain = 0:8192,
        xtick = {0,1024,...,8192},
        xmin = 0,
        xmax = 8192,
        legend pos = outer north east,
        cycle list name = exotic
      ]
      %\draw [<->, line width=0.8, color=red] (axis cs:0,8e5) -- (axis cs:4096,8e5) ;
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
      %\addplot table[x expr = \thisrowno{0}*16, y = r0107:u ] \stackoffsettable ;
      \addlegendentry{cycles:u} ;
      \addlegendentry{r0107:u} ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

@@

\begin{figure}[float=h]
  \caption{Important sections of virtual memory on a 64 bit process}
  \label{fig:virtualmemory}
  \begin{tikzpicture}[font=\footnotesize]

@@

  \end{tikzpicture}
\end{figure}

@@

To illustrate how this can cause bias, we revisit the example first presented in~\cite{Mytkowicz:2009:WrongData}, shown in Listing~\ref{lst:loopkernel}. 
This example is interesting for several reasons. 
The bias effects are significant and easily reproducible, while the example code is simple and easy to analyze.
Still, no satisfactory explanation as to what causes bias was given in the original paper.

\begin{lstlisting}[float=h, language=C, caption={Microkernel succeptible to aliasing between static variables and automatic variables for certain environment sizes}, label={lst:loopkernel}, frame=lines]
static int i, j, k;
int main() {
    int g = 0, inc = 1;
    for (; g < 65536; g++) {
        i += inc;
        j += inc;
        k += inc; 
    }
    return 0;
}
\end{lstlisting}

Their program contains five variables, two of which are stack allocated and i, j and k that are statically allocated.
Performance counter measurements of cycle counts are shown for 512 different environment sizes in Figure~\ref{fig:envbias}.
Sampling is done for every 16 byte increment of environment size, covering two 4K periods of addresses.
A finer sampling is not necessary, because the stack is by default aligned to 16 byte.
The program is compiled with GCC using no optimization.
Any optimization would likely disregard most of the function as redundant code, and reduce it to return zero immediately.

There are clearly two worst cases, indicated by significant spikes around the 3100 and 7100 byte offsets.
We analyzed these cases by sampling an extensive set of performance counters and correlating them with cycle counts.
The results show a high number of resource stalls when the spikes occur, indicated by high values for performance counters RESOURCE\_STALLS.ANY, CYCLE\_ACTIVITY.CYCLES\_LDM\_PENDING, CYCLE\_ACTIVITY.CYCLE\_NO\_EXECUTE and RESOURCE\_STALLS.RS.
We also find that LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS correlates almost perfectly with cycle count, which can explain the high number of resource stalls with false dependency from aliasing.

A small amount of assembly code was used to calculate the addresses of each stack allocated variable and output them directly using system calls, without impacting run time addresses.
We find that the spike in cycle count occurs when the address of g and inc alias with i and j, respectively.
Resource stalls are generated because of false dependencies between the addresses of (g, i) and (inc, j), resulting in worse performance.
Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment.
The program is \emph{biased} towards environment sizes that avoids this specific stack alignment.

\subsection{Heap Address Aliasing}
Address aliasing can be caused by conflicting pairs of load/store operations to any part of memory.
The previous section looked specifically at an example of collision between stack variables and static data.
Another scenario to consider is collisions in heap allocated memory.
In particular, code that operate on pairs of contiguous arrays can be vulnerable to 4K aliasing.
Acquiring dynamic memory at run time is usually done by calling malloc, which takes a number of bytes to allocate and returns a pointer to that area.
A typical malloc implementation uses two different strategies for how to allocate memory.
The \emph{heap} area in virtual memory is used for smaller allocations.
Programs are initially given a small heap area, which the allocator can increase by calling brk or sbrk.
For larger allocations, it is typically faster to use anonymous memory mapping by calling mmap, instead of growing the ``normal'' heap.
The heap starts at a low address close to the data section and grows towards higher addresses. 
Mmap allocates much closer to the stack.
While heap addresses can look like 0x16e30a0 or 0x1723020, pointers returned by mmap can for example be 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, mmap has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, thus every mmap'ed address ends with 0x010}

Consider the function shown in Listing~\ref{lst:conv}.
It computes the convolution between an input array and a fixed five-element kernel, writing the result to another array.
For simplicity, endpoints are not handled.

\pgfplotstableread{bin/convolution.dat}{\convolutiontable}
\begin{figure}[t]
  \caption{Cycle and alias counts for different offsets between input and output arrays in Listing~\ref{lst:conv}. Offset 0 means equal 12 bit address suffix, which default behavior for mmap allocations. Array size $n=2^{20}$, compiled with -O3. }
  \label{fig:heapalias}
  \begin{tikzpicture}
    \begin{axis}[
        font = \small,
        xlabel=Offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convolutiontable ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convolutiontable ;
      \addlegendentry{cycles:u} ;
      \addlegendentry{r0107:u} ;
    \end{axis}
  \end{tikzpicture}
\end{figure}

\begin{lstlisting}[float=h, language=C, caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.}, label={lst:conv}, frame=lines]
static float k[5]={0.1,0.25,0.3,0.25,0.1};
void conv(int n, float *in, float *out)
{
  int i, j;
  for (i = 2; i < n-2; ++i) {
    out[i] = 0;
    for (j = 0; j < 5; ++j)
      out[i] += in[i-2+j] * k[j];
  }
}
\end{lstlisting}

Figure \ref{fig:heapalias} shows ...

Using malloc to allocate input and output float arrays of size n, they will alias if mmap is used internally.
For sufficiently large values of n, this will always be the case.
Compiling with optimization O3 and n=$2^{20}$, a very large number of alias events and resource stalls are generated compared to cycle count.

@@

Some effort has been made to model properties of the LSD on previous Intel architectures, as an aid in low-level optimization. 
Hundt et. al.~\cite{Hundt:2011:MAO} looks at assembly rewrite rules for optimizing agains the Core 2 LSD.
For loops to be recognized, they have to fit within a set of restrictions imposed by the hardware implementation that tries to detect them. 
For instance, because the LSD cache is of fixed size there will be a maximum number of instructions a loop can contain for it to be recognized.
The capabilities of the LSD has improved with newer micro-architectures.
Some important limitations for Haswell's LSD are listed below.
\begin{itemize}
  \item Up to 12 chunk fetches of 32 B instruction memory.\footnote{The optimization manual actually states 8 chunk fetches for HTT enabled, and 11 for HTT disabled. Our experiments with a minimal test case shows that the LSD is active for loops spanning up to 12 chunks -- regardless of hyper threading.}
  \item Up to 28 micro-operations for HTT enabled, 56 for HTT disabled.
  \item Can not contain CALL or RET instructions
  \item Stack operations must match, i.e push must be matched with a pop.
\end{itemize}

@@

%\onecolumn
